\\Large The heat is on: Rising temperatures alter how and where mammals move

### 2025-08-29 writing/manuscript.Rmd
started cleaning up intro

* @gerlich_species_2025: effects of temperature on fly movement in the arctic
* "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C" (https://www.sciencedirect.com/science/article/pii/S2214581823000770#fig0015)
* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495

@@

<!-- movement is a convenient way to respond to change, but movement depends on temperature -->

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016].

Temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], altered phenology [@bunnell_global_2011; ], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and extreme weather [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022; @berger_climate_2018]. Changes in temperature thus affect how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

<!-- since movement depends on temperature, the consequences of movement do, too -->

By impacting animal movement behavior, temperature also affects the consequences of movement, such as encounter rates with resources [@mezzini_how_2025], other animals [@martinez-garcia_how_2020], and humans [@weststrate_how_2024], and these changes are exacerbated at extreme temperatures [@wilson_stenotherms_2001; @berger_climate_2018; @giroux_activity_2023]. But what counts as "extreme" varies across species: Species that experience narrow temperature ranges tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while thermal generalists can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. However, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019], affect many factors [e.g., temperature, phenology, and resource availability: @polazzo_measuring_2024], or are too rapid to respond within the span of a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

>>>>

* @gerlich_species_2025: effects of temperature on fly movement in the arctic
* "Potential changes in climate indices in Alberta under projected global warming of 1.5–5 $^\circ$C" (https://www.sciencedirect.com/science/article/pii/S2214581823000770#fig0015)
* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
* marine turtles shift and lose habitat with temperature: https://www.science.org/doi/10.1126/sciadv.adw4495
* changes in temperature change phelogy and movement behavior as a consequence @gerlich_species_2025
* @walker_global_2019: Global climate change and invariable photoperiods: {A} mismatch that jeopardizes animal fitness
\clearpage

@@

<!-- movement is a convenient way for ANIMALS to respond to change, but movement depends on temperature -->

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to optimize resource use [e.g., food, water, shelter$;$ @charnov_optimal_1976; @kacelnik_ideal_1992; @mezzini_how_2025], optimize habitat use [@winter_forecasting_2024; @schwerdt_ecological_2024], and avoid predators and competitors [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024; @torrez-herrera_monkeys_2020]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022] and the risk of hyperthermia [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [@hall_microrefuges_2016]. By altering mammals' movement speed and habitat use [jointly, their movement behavior: @nathan_movement_2008; @nathan_big-data_2022], ambient temperature also affects the consequences of movement behavior, such as encounter rates with resources [@mezzini_how_2025], humans [@weststrate_how_2024], predators, prey, or competitors [@martinez-garcia_how_2020; @glass_trade-off_2021; @brivio_seeking_2024]. Behavioral changes to temperature are exacerbated in extreme cold [@wilson_stenotherms_2001; @berger_climate_2018] and extreme heat [@powers_hovering_2017; @alston_temperature_2020; @giroux_activity_2023; @verzuh_behavioural_2023].

<!-- mammals  -->

Mammals are particularly susceptible to adverse effects from excessive heat [@sherwood_adaptability_2010]. While extreme cold often leads to reduced movement, provided that individuals can find refuge and take advantage of their body heat [@berger_climate_2018; @hou_cold_2020; @glass_trade-off_2021], extreme heat can often result in more rapid and serious physiological damage [@jessen_temperature_2001; @sherwood_adaptability_2010; @mota-rojas_physiological_2021]. Due to mammals' difficulty at dissipating heat, particularly when ambient temperatures are near or above their body temperature, they are particularly sensitive to extreme heat, particularly if they have a large body size [@dyer_travel_2023; @verzuh_behavioural_2023]. Furthermore, species that experience narrow temperature ranges (i.e., $`$temperature specialists') tend to be more sensitive to changes [e.g., giant anteaters, *Myrmecophaga tridactyla*: @giroux_activity_2023], while species that experience wide ranges of temperature variation (i.e., $`$temperature genralists') can adapt more easily to change [@levins_evolution_1974; @botero_evolutionary_2015; @dupont_beyond_2024]. Still, generalists' greater adaptability is tested when they are subject to changes that are more severe [@mccain_assessing_2019; @rantanen_arctic_2022], affect many factors at once [e.g., temperature, snow depth, ecosystem phenology, and resource availability: @polazzo_measuring_2024], or are too rapid and severe to respond within a lifetime or a few generations [@levins_evolution_1974; @fuller_towards_2016; @xu_plasticity_2021].

@@

***tie to text above:***

***...*** Temperature also has indirect effects on movement through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010], snow depth and density [@montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], altered phenology [@bunnell_global_2011; ], and the frequency and intensity of extreme events [@zhang_increased_2023; @bunnell_global_2011], including forest fires, droughts, and extreme weather [@zurowski_summer_2023; @duncan_predicting_2012; @bunnell_global_2011; @merkle_large_2016; @slatyer_ecological_2022; @berger_climate_2018]. Changes in temperature thus affect how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].
